### 2013-10-16 chi2014/qualitative.tex
revision with stats

In addition to modelling data and confirming our hypotheses, we performed a qualitative analysis. The primary goal was to identify trends  that were repeated between participants. These trends can then be used to inform and justify interventions to change the types of ideas produced.

@@

A \emph{strategy} is a plan, rule, or idea that is used by a participant to generate brainstorming responses.
Without debriefing participants, we cannot infer the goal of a strategy, but define it in terms of its observable effects.
These effects are repeated, cross-participant presence of classes of similar ideas.
Not all participants utilized all strategies in their brainstorming runs.

We identified X \emph{orthogonal} strategy categories - each category of strategies can be employed by a participant with or without any of the other categories.
The first, \emph{problem scoping}, is a process in which a participant chooses a limited implication or component of the problem and provides multiple specific solutions.
The second category expands on the \emph{riffing} phenomenon introduced earlier, identifying X different types of riffing exployed.
Finally, \emph{language use} strategies... TODO
%Finally, the \emph{humour} strategy emphasizes generating ideas that are amusing rather than practical.

@@

To complete the brainstorming task, participants had to generate some number of uses for a partially broken MP3 player. Under the problem scoping strategy, this problem can reduced in scope to one of many sub-problems: brainstorm uses for \emph{an audio output device}; brainstorm uses for \emph{an electronic screen}; brainstorm uses for \emph{a hard drive}; and so on. The participant chooses one sub-problem at a time and generates solutions, each of which is also a solution to the original problem. For example, the following are generated by a participant focussing on the audio output capabilities of an MP3 player:

@@

The previous examples of problem scoping are examples of emphasizing specific details of the problem. Problem scoping can also entail neglecting details. For example, the following ideas from a participant apply to almost any physical object, not just the MP3 player described:

@@

We refer to the former, detail-oriented kind of problem scoping as \emph{scoping inward} and the alternative as \emph{scoping outward}. Problem scoping is employed by all participants, but the type of scoping (inwards vs outward) is a useful discriminant.

@@

\emph{Generalization riffing} occurs when a participant generates two or more ideas and one is a generalization of the other. For example, two consecutive ideas given by a participant:

@@

Another type of riffing is \emph{repeat riffing}. Under this strategy, a participant produces multiple reponses with only slight differences between them. For example:

\begin{itemize}
    \item recycle to make better MP3/ipod players
    \item use the circuits to for the ipod/MP3 to power other things that are more recent
\end{itemize}

@@

\emph{Hold riffing} is a strategy in which participants hold at least one element of a previous response constant when generating a new response, such that a new idea is formed. This can be thought of as analogous to the game of \emph{draw poker}, in which players hold some cards while receiving new ones, with the intent of building a winning hand complementing the held cards.
For example:

@@

Hold riffing can also involve holding language terms constant even when the meaning of the term changes between responses, as in free association. For example:

@@

\emph{Continuation riffing} is the strategy of creating another idea that assumes the previous idea has already been implemented. The new ideas relies on the implications or consequences of the previous. For example:

@@

Within and across these strategies, we identified two sub-strategies or tweaked behaviours.
Some participants \emph{pair} their riffs. These participants  often riff on old ideas but never produce more than two versions of each idea. For example, paired hold riffing from a participant:

\begin{itemize}
\item TODO
\item TODO
\end{itemize}

Finally, \emph{reaching back} is a variant of riffing employed by almost all participants who provided more than 10 responses. Reaching back entails riffing on an idea that occured further back in the run than the previous response. The presence of reaching back speaks to the exhaustion of the participant's creativity, as they produce new variants of old ideas rather than new ideas altogether.

Almost all participants employ some form of riffing. This is not surprising, as it is ingrained in the rules of brainstorming given at the beginning of the task: "combinations of ideas count as new ideas".

@@

Participants varied in the language they used to express their ideas. Three strategies were observed: language repitition, brevity, and ambiguity.

Some participants re-use sentence structure or phrasings accross multiple responses. We call this strategy \emph{language repitition}. For example:

@@

In this example, while language is re-used, each instance is a new idea.
However, language repitition is especially common in repeat riffing, as this example shows:

@@

\emph{Brevity} is the simple strategy of producing short responses. Answers under this strategy are not neccessarily less specific than longer answers. For example "textbook audio reader" is a concise description of a very specific use for an MP3 player.

@@

Above, we described the problem scoping strategy. Some participants performed \emph{problem scoping without solution}, a variant of this strategy in which problem scoping is the only component of the response.
For example, one participant's response to the MP3 player problem was to "Remove glass screen to make something". In this case, the participant has provided a scoping of the problem (focus on the glass component) but has not actually provided a solution to the scoped problem.

When a solution is provided, it is still occasionally lacking in sufficient information to act upon. We call this strategy \emph{goal without plan}. For example, a common response to the problem was "use as art". In this case, an end goal has been established, but any implementation of the goal requires further idea generation. The inverse of this strategy is \emph{plan without goal}, as in the response "change the settings". In this case, one could certainly perform the operation on the MP3 player, but it doesn't provide any obvious benefit.

In many cases, responses that do not have an explicit goal or plan nonetheless imply one. For example, "throw it in the ocean" is an action performed with the MP3 player with no stated goal, but the implied goal is to get rid of the broken device. In these cases, as judged by the coder, reponses were not considered part of a partial solution strategy.

Finally, responses in the \emph{pass the problem} strategy simply relocate the need for idea generation to a third party. Common responses under this strategy are "give as gift" or "donate to cancer research". These solve the original problem, but simultaneously re-open the problem under slightly changed conditions. In this case, some cancer researchers must determine what to do with broken MP3 players.

\subsection{Personas}

A persona can be thought of as a toolbox of strategies that we see repeated across participants. Participants of a given persona have a subset of strategies that are used often, and a subset that are rarely used. In a process analogous to affinity diagramming, brainstorming runs were clustered based on similarity of strategy use. This resulted in six clusters of repeated behaviour, each of which is a basis of a persona. These personas are summarized in Table TAB.

The \emph{high concept} persona consists of participants who view the MP3 player as an abstract object and provide many distinct, if often vague, responses. Commonly employed strategies are scoping outward, hold riffing, and partial responses.

\emph{Outside the box} personas are similar to high concept, except that their responses are more well-defined. Common strategies are scoping outward, hold riffing and brevity. Partial responses are rare.

and so on...

\subsubsection{Outliers}

In addition to the observed personas, there were several outliers. One participant employed no strategies to a signficant degree (in the case of problem scoping, they scoped inward and outward with similar frequency). A second outlier responded with questions, such as "What can the average person do to better the environment?".

\begin{table*}
    \begin{tabular}{|l|p{0.35\linewidth}|l|}
        \hline
        \textbf{Persona} & \textbf{Common strategies} & \textbf{Rare strategies} \\
        \hline
        High concept & scoping outward, hold riffing, partial responses & brevity, reach-backs \\
        \hline
        Outside the box & scoping outward, hold riffing, brevity & partial responses, reach-backs \\
        \hline
        broad maximizers & repeat riffing, brevity, \newline
        partial responses  & reach-backs \\ 
        \hline
        broad explorers & hold riffing, partial responses & \\
        \hline
        kitchen sink & scoping outward, repeat riffing, brevity, \newline
        partial responses, reach-backs & \\ 
        \hline
        detail explorers & scoping inward, hold riffing & brevity, partial responses, reach-backs\\
        \hline
    \end{tabular}
    \caption{Strategies used by personas}
\end{table*}

>>>>

In addition to modeling data and confirming our hypotheses, we performed a qualitative analysis. The primary goal was to identify trends  that were repeated between participants. These trends can then be used to inform and justify interventions to change the types of ideas produced.

@@

A \emph{strategy} is a plan, rule, or idea that is used by a participant to generate brainstorming responses.
Without debriefing participants, we cannot infer the goal of a strategy, but define it in terms of its observable effects.
These effects are observed as repeated, cross-participant classes of similar ideas.
Not all participants utilized all strategies in their brainstorming runs.

We identified 3 \emph{orthogonal} strategy categories - each category of strategies can be employed by a participant with or without any of the other categories.
The first, \emph{problem scoping}, is a process in which a participant chooses a limited implication or component of the problem and provides multiple specific solutions.
The second category expands on the \emph{riffing} phenomenon introduced earlier, identifying 4 different types of riffing explored.
Finally, \emph{language use} strategies relate not to the meaning of the ideas generated, but to the way they are expressed.
%Finally, the \emph{humour} strategy emphasizes generating ideas that are amusing rather than practical.

@@

To complete the brainstorming task, participants had to generate some number of uses for a partially broken MP3 player. Under the problem scoping strategy, this problem can reduced in scope to one of many sub-problems: brainstorm uses for \emph{an audio output device}; brainstorm uses for \emph{an electronic screen}; brainstorm uses for \emph{a hard drive}; and so on. The participant chooses one sub-problem at a time and generates solutions, each of which is also a solution to the original problem. For example, the following are generated by PXXX focusing on the audio output capabilities of an MP3 player:

@@

The previous examples of problem scoping are examples of emphasizing specific details of the problem. Problem scoping can also entail neglecting details. For example, the following ideas from PXXX apply to almost any physical object, not just the MP3 player described:

@@

We refer to the former, detail-oriented kind of problem scoping as \emph{scoping inward} and the alternative as \emph{scoping outward}. These two types of scoping cover almost all responses. However, some responses employ no scoping whatsoever. These responses utilize the device exactly as it was intended, as a portable MP3 player. For example, "get it fixed and give it to a needy kid".

Problem scoping is employed by all participants, but the type of scoping is a useful discriminant.

@@

\emph{Generalization riffing} occurs when a participant generates two or more ideas and one is a generalization of the other. For example, two consecutive ideas given by PXXX:

@@

Another type of riffing is \emph{repeat riffing}. Under this strategy, a participant produces multiple responses with only slight differences between them. For example (PXXX):

\begin{itemize}
    \item recycle to make better MP3/iPod players
    \item use the circuits to for the iPod/MP3 to power other things that are more recent
\end{itemize}

@@

\emph{Hold riffing} is a strategy in which participants hold at least one element of a previous response constant when generating a new response, such that a new idea is formed. For example (PXXX):

@@

Hold riffing can also involve holding language terms constant even when the meaning of the term changes between responses, as in free association. For example (PXXX):

@@

\emph{Continuation riffing} is the strategy of creating another idea that assumes the previous idea has already been implemented. The new idea on the implications or consequences of the previous. For example (PXXX):

@@

\paragraph{Spatial separation}

The strategies above describe techniques for taking one idea and transforming it into another. In this section, we briefly describe patterns of \emph{where} participants implemented these strategies.

The most natural separation between a riffed idea and its source is none - the riffed idea directly follows the source. We call this \emph{linear riffing}. Some participants \emph{pair} their riffs. Paired riffing is a special case on linear riffing in which only a single riffed idea is produced, and no further riffs are produced on that source for the remainder of the run. For example, paired repeat riffing from PXXX:

\begin{itemize}
\item cut up and use to decorate shoes
\item cut up and use to decorate vase
\end{itemize}

Finally, \emph{reach backs} comprise all other riffs. reaching back entails riffing on an idea that occurred further back in the run than the previous response.

@@

Some participants re-use sentence structure or phrasings across multiple responses. We call this strategy \emph{language repetition}. For example (PXXX):

@@

In this example, while language is re-used, each instance is also a new idea unrelated to the previous.
However, language repetition is especially common in repeat riffing, as this example (PXXX) shows:

@@

@@

Above, we described the problem scoping strategy. Some participants performed \emph{problem scoping without solution}, a variant of this strategy in which problem scoping is the only component of the response.
For example, on of PXXX's responses to the MP3 player problem was to "Remove glass screen to make something". In this case, the participant has provided a scoping of the problem (focus on the glass component) but has not actually provided a solution to the scoped problem.

In some responses, an end goal has been established, but any implementation of the goal requires further idea generation. We call this strategy \emph{goal without plan}. For example, a selection of responses from PXXX:

\begin{itemize}
    \item use old parts to make a new device
    \item old parts can use to make something
    \item create a new device
    \item reuse all working parts
\end{itemize}

In this case, the goal of creating a new electronic device is given, but it is not clear what will be created or how the materials will be re-used.

The inverse of this strategy is \emph{plan without goal}, as demonstrated in these responses (also from PXXX):

\begin{itemize}
    \item melt down old parts
    \item see what old parts are usable
\end{itemize}

In this case, the operations of melting down the device or checking for working parts can be performed, but don't provide any obvious benefit.

In many cases, responses that do not have an explicit goal or plan nonetheless imply one. For example, "throw it in the ocean" (PXXX) is an action performed with the MP3 player with no stated goal, but the implied goal is to get rid of the broken device. In these cases, as judged by the coder, responses were not assumed to employ a partial solution strategy.

Finally, responses in the \emph{pass the problem} strategy simply relocate the need for idea generation to a third party. These responses from PXXX provide an example:

\begin{itemize}
    \item donate to less fortunate that are trying to assist the homeless
    \item give to university
    \item give to non-profit that can benefit
    \item give to an organization that has the knowledge to use these devices
\end{itemize}

These solve the original problem, but simultaneously re-open the problem under slightly changed conditions. In this case, a university/non-profit/other body must determine what to do with broken MP3 players.

\subsection{Strategy use}

In this section, we briefly describe how strategies were employed by participants. The primary author again coded 30 randomly selected runs (5 from each condition, 1150 instances total) for the strategies identified above. Note that examples of riffing identified in this section are distinct from the algorithmic/quantitative identification of riffing described in the method section above. The prevalence of each strategy as a function of total number of instances is given in Table TAB.

\begin{table}
    \begin{tabular}{r | l l}
        \textbf{strategy} & \textbf{\# instances} & \textbf{\% of instances} \\
        inward scoping & 267 & 23.2 \\
        outward scoping & 828 & 72 \\
        no scoping & 55 & 4.8 \\
        \hline \hline
        hold riffing & 278 & 24.2 \\
        continuation riffing & 3 & 0.3 \\
        repeat riffing & 119 & 10.3 \\
        generalization riffing & 48 & 4.2 \\
        \hline
        any riffing & 448 & 39.0 \\
        pair & 108 & 9.4 \\
        reach back & 129 & 11.2 \\
        \hline \hline
        scoping without solution & 3 & 0.3 \\
        passing the problem & 69 & 6 \\
        plan without goal & 58 & 5.0 \\
        goal without plan & 58 & 5.0 \\ 
        \hline
        any partial solution & 188 & 16.3 \\
    \end{tabular}
\end{table}

Outward scoping is by far the most commons strategy, occurring in 72\% of instances. Riffing is also very common, with 40\% of instances employing some form of riffing. Riffs are most often consecutive (linear riffing), and hold riffing is the most common type.

Partial solutions are fairly uncommon (16.3\%), and most often involve passing the problem (6\%). Continuation riffing and scoping without solution are the least common strategies, each with only 3 instances.

\subsubsection{Strategy location}

We also consider where in a brainstorming run the strategies are more likely to occur. The results are summarized in Table TAB.

\begin{table}
    \begin{tabular}{r | l}
        \textbf{strategy} & \textbf{mean position in run} \\
        inward scoping & 26.2 \\
        outward scoping & 34.8 \\
        no scoping & 18.1 \\
        \hline \hline
        hold riffing & 34.9 \\
        continuation riffing & 16.3 \\
        repeat riffing & 40.3 \\
        generalization riffing & 26.4 \\
        \hline
        pair & 31.9 \\
        reach back & 33.1 \\
        \hline \hline
        scoping without solution & 39.7 \\
        passing the problem & 40 \\
        plan without goal & 29.6 \\
        goal without plan & 20.0 \\ 
    \end{tabular}
\end{table}

Most of the differences run position between strategies are minimal. Most discriminatory are the differences in scoping. Lack of scoping is, on average, very early in the run, while scoping inward towards some detail of the problem occurs later. Outward scoping, or dropping details of the problem, occurs latest of all. This gradual increase hints that more "outside the box" ideas may not show up until later in runs, after participants have finished exploring specific details of the problem.

\subsubsection{Strategy co-occurrence}

Of final interest was the question of which strategies occurred together. For each run, a count was taken of the number of occurrences of each strategy, and the Pearson r of these counts was taken for each pairing of strategies to give a rough measure of how the strategies were grouped in runs. Strong correlations (greater than 0.6) are given in Table TAB.

\begin{table}
    \begin{tabular}{l l l}
        \textbf{strategy 1} & \textbf{strategy 2} & \textbf{Pearson r} \\
        \hline
        passing the problem & repeat riffing & 0.89 \\
        hold riffing & outward scoping & 0.85 \\
        pairing & scoping without solution & 0.64 \\
        reaching back & hold riffing & 0.64 \\
        pairing & outward scoping & 0.63 \\
        reaching back & outward scoping & 0.62 \\
        plan without goal & hold riffing & 0.61 \\
    \end{tabular}
\end{table}

Passing the problem and repeat riffing were highly correlated (0.89), indicating that responses utilizing pass the problem were likely to be repeated many times to generate more responses. For each of the other correlations, the potential implications less obvious. Further analysis is required to tease out the implications of strategy co-occurrence.
